---
title: "2024 KPMP Pseudobulk Files "
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: false
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo=T, error=FALSE, message=FALSE, warning=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("SeuratDisk")) {install.packages("SeuratDisk"); require("SeuratDisk")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("paletteer")) {install.packages("paletteer"); require("paletteer")} # color palette
if (!require("grDevices")) {install.packages("grDevices"); require("grDevices")} # for grDevices palette
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("tibble")) {install.packages("tibble"); require("tibble")} # for table transformation
if (!require("geneName")) {install.packages("geneName"); require("geneName")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("ggupset")) {install.packages("ggupset"); require("ggupset")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("viridis")) {install.packages("viridis"); require("viridis")}

library(SeuratData)
library(openxlsx)
library(gplots)
library(ggvenn)

```

# Introduction

This file was made to create pseudobulk matrix files from the 2024 KPMP snRNAseq data release in order to identify DEGs that are related to PFAS exposure. 

The Pseudobulk and cell count files are made at 3 different levels of annotation: <br>
1) Class <br>
2) Subclass.l1 <br>
3) Subclass.l2 <br>

Working with the KPMP snRNAseq file: **WashU-Altos_KPMP-Biopsy_10X-R_01082024.h5Seurat**

# 2024 snRNASeq KPMP Object

## Load KPMP object (2024 data file)

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Sys.time()
KPMP <- LoadH5Seurat(here("WashU-Altos_KPMP-Biopsy_10X-R_01082024.h5Seurat")) 
Sys.time()

head(KPMP@meta.data)

DimPlot(KPMP, group.by = "subclass.l1", label = TRUE)

KPMP

```

## Add Clinical information into meta.data

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

meta <- read.csv(here("OpenAccessClinicalData.csv"))

df <- FetchData(KPMP, "patient") %>% rownames_to_column(var = "CellID")

df <- left_join(df, meta, by = c("patient" = "Participant.ID")) %>% column_to_rownames(var = "CellID")

KPMP <- AddMetaData(KPMP, df)

head(KPMP@meta.data)

unique(KPMP@meta.data$Enrollment.Category)

KPMP@meta.data$Enrollment.Category <- factor(KPMP@meta.data$Enrollment.Category, levels = c("Healthy Reference", "AKI", "CKD", "DM-R"))

```

# Create Files

## Subclass.l1

### Pseudobulk Matrix

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(KPMP, group.by = "subclass.l1", label = TRUE)

ShotgunPB <- AggregateExpression(KPMP, 
                           group.by = c("subclass.l1", "patient"), 
                           assays = 'RNA', 
                           slot = "counts",
                           return.seurat = FALSE)

ShotgunPB <- ShotgunPB$RNA
ShotgunPB <- as.matrix(ShotgunPB)  # Convert to matrix
ShotgunPB <- t(ShotgunPB) # Transpose matrix
ShotgunPB <- as.data.frame(ShotgunPB) # Convert from matrix to dataframe 

ShotgunPB_20 <- ShotgunPB[, 1:20]

head(ShotgunPB_20, 15)

write.csv(ShotgunPB, here("2024KPMP_Psuedobulk_Subclassl1.csv"))

```

### Cell Number

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Idents(KPMP) <- KPMP@meta.data$subclass.l1

t1 <- table(Idents(KPMP), KPMP@meta.data$patient)

t1

t1 <- as.matrix(t1)
t2 <- as.data.frame(t1) 

t2

write.csv(t2, here("2024KPMP_CellNumber_Subclassl1.csv"))

```

## Subclass.l2

### Pseudobulk Matrix

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(KPMP, group.by = "subclass.l2", label = TRUE)

ShotgunPB <- AggregateExpression(KPMP, 
                           group.by = c("subclass.l2", "patient"), 
                           assays = 'RNA', 
                           slot = "counts",
                           return.seurat = FALSE)

ShotgunPB <- ShotgunPB$RNA
ShotgunPB <- as.matrix(ShotgunPB)  # Convert to matrix
ShotgunPB <- t(ShotgunPB) # Transpose matrix
ShotgunPB <- as.data.frame(ShotgunPB) # Convert from matrix to dataframe 

ShotgunPB_20 <- ShotgunPB[, 1:20]

head(ShotgunPB_20, 15)

write.csv(ShotgunPB, here("2024KPMP_Psuedobulk_Subclassl2.csv"))

```

### Cell Number

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Idents(KPMP) <- KPMP@meta.data$subclass.l2

t1 <- table(Idents(KPMP), KPMP@meta.data$patient)

t1

t1 <- as.matrix(t1)
t2 <- as.data.frame(t1) 

t2

write.csv(t2, here("2024KPMP_CellNumber_Subclassl2.csv"))

```

## Class

### Pseudobulk Matrix

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(KPMP, group.by = "class", label = TRUE)

ShotgunPB <- AggregateExpression(KPMP, 
                           group.by = c("class", "patient"), 
                           assays = 'RNA', 
                           slot = "counts",
                           return.seurat = FALSE)

ShotgunPB <- ShotgunPB$RNA
ShotgunPB <- as.matrix(ShotgunPB)  # Convert to matrix
ShotgunPB <- t(ShotgunPB) # Transpose matrix
ShotgunPB <- as.data.frame(ShotgunPB) # Convert from matrix to dataframe 

ShotgunPB_20 <- ShotgunPB[, 1:20]

head(ShotgunPB_20, 15)

write.csv(ShotgunPB, here("2024KPMP_Psuedobulk_class.csv"))

```

### Cell Number

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Idents(KPMP) <- KPMP@meta.data$class

t1 <- table(Idents(KPMP), KPMP@meta.data$patient)

t1

t1 <- as.matrix(t1)
t2 <- as.data.frame(t1) 

t2

write.csv(t2, here("2024KPMP_CellNumber_class.csv"))

```


# Session Info

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Sys.time()

sessionInfo()

```
